# Findings

Figure \@ref(fig:ecdf) shows the empirical cumulative density function for the 
headway deviation data, grouped by TSP threshold. Note that a "perfect"
headway distribution, where all vehicles maintain the exactly intended headway,
would be a step function with a vertical transition at 6 minutes. 
Visually, the difference  between the various threshold settings is not dramatic. 
The 2-minute threshold appears to have slightly more vehicles arrive behind the
scheduled headway (6 minutes), and slightly more arrive before it, than the
other two threshold groups. The median of the distribution for all
four thresholds is remarkably similar and is just a few seconds behind the target
headway. This observation strengthens our determination to examine the sides of
the distribution --- rather than its center --- with a quantile regression model.

```{r ecdf, dev=if(knitr:::is_latex_output()) 'tikz' else 'svg', fig.cap="Cumulative probability distribution of headway deviation by threshold."}
tar_load(ecdf)
ecdf + 
  xlab("Actual Headway [minutes]") +
  ylab("Cumulative Probability") +
  scale_color_discrete("TSP Threshold") + 
  coord_cartesian(xlim = c(0, 12)) +
  scale_y_continuous(breaks = c(0, 0.1, 0.15, 0.5, 0.85, 0.9, 1)) +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12))
```


```{r models}
tar_load(models)

coef_map <- c(
  "(Intercept)" = "(Intercept)", 
  "threshold2 min" =  "TSP: 2 minutes",
  "thresholdAlways" = "TSP: Always",
  "directionSB"   = "Southbound", 
  "periodAM Peak" = "AM Peak", 
  "periodPM Peak" = "PM Peak", 
  "cumdwell" = "Cumulative Dwell [minutes]", 
  "directionSB:periodAM Peak" = "Southbound $\\times$ AM Peak",  
  "directionSB:periodPM Peak" = "Southbound $\\times$ PM Peak"
)

f1 <- function(x) format(round(x, 3), big.mark=",")
gm <- list(
  list("raw" = "AIC", "clean" = "AIC", "fmt" = f1),
  list("raw" = "logLik", "clean" = "Log Likelihood", "fmt" = f1)
)
```

Models estimating the headway effects of TSP requesting threshold and other factors 
at an array of distribution quantiles are given in Table \@ref(tab:models-summ).
In these models, the "(Intercept)" represents the expected headway at that percentile
before considering additional information. For example, the 10th-percentile headway
is `r coef(models[["0.1"]]$All)[1]` minutes, all else equal. The other
coefficients in the models modify this average headway. Almost all coefficients are 
significant, and many serve to widen the headway distribution.
Buses traveling in the PM peak, for example, have a significantly higher 
90th-percentile headway and a lower 10th-percentile headway than buses
traveling outside of the AM or PM peaks. Buses traveling in the
southbound direction also have a wider headway distribution than those in the 
northbound direction (implied to be $0$), and each additional minute of cumulative dwell
time widens the headway distribution.

```{r models-summ}
modelsummary(
  lapply(models, function(x) x$All) %>% 
    set_names(str_c(as.numeric(names(models)) * 100, "th")),
  coef_map = coef_map, gof_map = gm,
  estimate = "{estimate} ({statistic}){stars}",
  statistic = NULL, title = "Quantile Regression Estimates", 
  notes = c("t-statistics in parentheses, * p < 0.1, ** p < 0.05, *** p < 0.01",
            "Coefficients represent change to expected headway in minutes."),
  escape=FALSE
  ) %>%
   kable_styling(latex_options = c("scale_down")) 
```

In contrast, the estimates reveal that TSP significantly 
*narrows* the expected headway distribution, with fewer long headways and fewer
short headways. And while most of the other explanatory variables have 
an effect on the median headway, TSP has no significant effect after these other 
variables have been controlled for. 
A potentially curious finding is that implementing a 2-minute TSP request
threshold improves headway adherence more than allowing every transit vehicle to
request TSP. This finding echoes the schedule-based TSP analysis by @sheffieldUDOT.

The schedule of threshold changes was not randomized in any way,
and it is possible that the results of this study are tied up in unaccounted
seasonal variation, or other omitted explanatory variables. These limitations
notwithstanding, we find that --- all else equal --- schedule-based TSP
marginally improves the headway adherence of UVX.
